StackTokens: Enforcing Well-bracketed Control Flow and Stack Encapsulation using Linear Capabilities

### 2018-07-01 fully-abstract-comp/popl2019/paper.tex
Some work on explaining the remaining problems with stack and return capabilities.

% %% Some recommended packages.
% \usepackage{booktabs}   %% For formal tables:
%                         %% http://ctan.org/pkg/booktabs
% \usepackage{subcaption} %% For complex figures with subfigures/subcaptions
%                         %% http://ctan.org/pkg/subcaption

@@

Well-bracketed control flow expresses that invoked functions must either return to their callers, invoke other functions themselves or diverge, and generally holds in programming languages that do not offer a primitive form of continuations. 
At the assembly level, this is not so obvious, as invoked functions get direct access to return pointers, that they are supposed to jump to a single time, at the end of their execution, but there is no guarantee that untrusted assembly code respects this intended usage.
Particularly, a function may invoke return pointers from other stack frames than its own: either frames higher in the call stack or ones that no longer exist as they have already returned before.

Local state encapsulation is the guarantee that when a function invokes another function, its local variables will not have been modified when the invoked function returns.
At the assembly level, this property is also far from obvious.
The calling function's local variables are stored on the stack during the invocation, and functions are not supposed to touch stack frames other than their own.
However, untrusted assembly code is free to ignore this requirement and overwrite the local state of its caller or other stack frames.

@@

To enforce well-bracketed control flow and local state encapsulation on a capability machine, there are essentially two approaches.
A first approach uses separate stacks for distrusting components, and a central, trusted stack manager component that mediates cross-component invocations.
This idea has been applied in CheriBSD (an operating system built on CHERI)~\citep{watson_capability_2015}, but it is not without downsides.
First, it scales poorly to large amounts of distrusting components, because of the need to reserve separate stack space for all components.
Also, in the presence of higher-order values (e.g. function pointers, objects etc.), the stack manager needs to be able to decide which component a higher-order value belongs to, in order to provide it the right stack pointer upon invocation and it is not clear how this can be done efficiently in the presence of large amounts of components.
Finally, this approach does not allow passing stack references between components.

A more scalable approach retains a single stack that is shared between components.
Enforce well-bracketed control flow and stack frame encapsulation in this approach, requires a way to temporarily provide stack and return capabilities to an untrusted component and revoke them after it returns.
While capability revocation is expensive in general, some capability machines offer restricted forms of revocation that can be implemented efficiently.
For example, CHERI offers a form of \emph{local} capabilities that, briefly, can only be stored in registers or on the stack, but not in other parts of memory.
In previous work, we have demonstrated that by making the stack and return pointer local, and a number of security checks and measures, well-bracketedness and local stack frame encapsulation can be guaranteed~\citep{skorstengaard_reasoning_2017}.
However, a problem with this approach is that revoking the local stack and return capabilities on every boundary crossing requires clearing the entire unused part of the stack, a potentially expensive operation.

In this work, we propose and study StackTokens: an alternative calling convention that enforces well-bracketed control flow and local state encapsulation with a single shared stack.
Instead of CHERI's local capabilities, it builds on \emph{linear} capabilities; a new form of capabilities that has not been previously described in the published literature.\footnote{Although they are mentioned here and there by different people.} 
These capabilities are enforced by the hardware to be non-duplicable.
We propose to make stack and return pointers linear and require components to hand them out in cross-component invocations and to return them in returns.
The non-duplicability of linear capabilities, together with some security checks, then allows us to guarantee well-bracketed control flow and local stack frame encapsulation, without large overhead on boundary crossings and particularly no need for clearing large blocks of memory.

A second contribution of this work is the way in which we formulate these two properties, using a technique we call \emph{fully abstract overlay semantics}.
Formulations in previous work are either partial and not suitable for reasoning~\cite{abadi_control-flow_2005} or lacked evidence of generality~\cite{skorstengaard_reasoning_2017}.
Our new formulation starts from the premise that security results for a calling convention should be reusable as part of a larger proof of a secure compiler.
To accomodate this, we define a second operational semantics for our target language, which has a native well-bracketed stack and primitive ways to do calls and returns.
This well-behaved semantics natively guarantees well-bracketed control flow and local stack frame encapsulation for components using our calling convention.
As such, these components can be sure that they will only ever interact with well-behaved other components that respect our desirable properties.
To express security of our calling convention, we then show that considering the same components in the original semantics, does not give adversaries additional ways to interact with them. 
More formally, we show that mapping a component in the well-behaved semantics to the same component in the original semantics is fully abstract~\cite{abadi_protection_1999}, i.e. that components are indistinguishable to arbitrary adversaries in the well-behaved language iff they are indistinguishable to arbitrary adversaries in the original language.

@@

\section{Linear Stack and Return Capabilities}
\begin{itemize}
\item informally explain how an adversary may try to abuse stack and return pointers
\item informally explain how we prevent this using linear capabilities
\item use the tikz pictures from the PriSC presentation to explain all of this
\end{itemize}
\section{A Capability Machine with Sealing and Linear Capabilities}
\begin{itemize}
\item present our target language and its operational semantics (excerpts)
\item mention roughly what components look like
\end{itemize}
\section{Formulating Security with a Fully Abstract Overlay Semantics}
\begin{itemize}
\item present our source language, its operational semantics (excerpts)
\item mention our assumption of reasonability
\item present the full abstraction theorem.
\item sketch high-level structure of the proof
\end{itemize}
\section{Discussion}
\begin{itemize}
\item mention that tail calls are supported through xcall
\item explain how fully abstract overlay semantics could form one pass of a verified secure compiler.
\end{itemize}
\section{Related Work}
\begin{itemize}
\item CFI and friends
\item our ESOP 2018 paper \citep{skorstengaard_reasoning_2017}
\item undefined behavior and full abstraction \citep{juglaret_beyond_2016}
\end{itemize}
\bibliography{references}

>>>>

% %% Some recommended packages.
% \usepackage{booktabs}   %% For formal tables:
%                         %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{wrapfig}

\input{preamble-illu}

@@

Well-bracketed control flow (WBCF) expresses that invoked functions must either return to their callers, invoke other functions themselves or diverge, and generally holds in programming languages that do not offer a primitive form of continuations. 
At the assembly level, this is not so obvious, as invoked functions get direct access to return pointers, that they are supposed to jump to a single time, at the end of their execution, but there is no guarantee that untrusted assembly code respects this intended usage.
Particularly, a function may invoke return pointers from other stack frames than its own: either frames higher in the call stack or ones that no longer exist as they have already returned before.

Local state encapsulation (LSE) is the guarantee that when a function invokes another function, its local variables (saved on its stack frame) will not have been read or modified when the invoked function returns.
At the assembly level, this property is also far from obvious.
The calling function's local variables are stored on the stack during the invocation, and functions are not supposed to touch stack frames other than their own.
However, untrusted assembly code is free to ignore this requirement and read or overwrite the local state of other stack frames.

@@

To enforce WBCF and LSE on a capability machine, there are essentially two approaches.
A first approach uses separate stacks for distrusting components, and a central, trusted stack manager component that mediates cross-component invocations.
This idea has been applied in CheriBSD (an operating system built on CHERI)~\citep{watson_capability_2015}, but it is not without downsides.
First, it scales poorly to large amounts of distrusting components, because of the need to reserve separate stack space for all components.
Also, in the presence of higher-order values (e.g. function pointers, objects etc.), the stack manager needs to be able to decide which component a higher-order value belongs to, in order to provide it the right stack pointer upon invocation and it is not clear how this can be done efficiently in the presence of large amounts of components.
Finally, this approach does not allow passing stack references between components.

A more scalable approach retains a single stack that is shared between components.
Enforcing WBCF and LSE in this approach, requires a way to temporarily provide stack and return capabilities to an untrusted component and revoke them after it returns.
While capability revocation is expensive in general, some capability machines offer restricted forms of revocation that can be implemented efficiently.
For example, CHERI offers a form of \emph{local} capabilities that, briefly, can only be stored in registers or on the stack, but not in other parts of memory.
In previous work, we have demonstrated that by making the stack and return pointer local, and a number of security checks and measures, the two properties can be guaranteed~\citep{skorstengaard_reasoning_2017}.
However, a problem with this approach is that revoking the local stack and return capabilities on every boundary crossing requires clearing the entire unused part of the stack, a potentially expensive operation.

In this work, we propose and study StackTokens: an alternative calling convention that enforces WBCF and LSE with a single shared stack.
Instead of CHERI's local capabilities, it builds on \emph{linear} capabilities; a new form of capabilities that has not been previously described in the published literature.\footnote{Although they are mentioned here and there by different people.} 
These capabilities are enforced by the hardware to be non-duplicable.
We propose to make stack and return pointers linear and require components to hand them out in cross-component invocations and to return them in returns.
The non-duplicability of linear capabilities, together with some security checks, then allows us to guarantee WBCF and LSE, without large overhead on boundary crossings and particularly no need for clearing large blocks of memory.

A second contribution of this work is the way in which we formulate these two properties, using a technique we call \emph{fully abstract overlay semantics}.
Formulations in previous work are either partial and not suitable for reasoning~\cite{abadi_control-flow_2005} or lacked evidence of generality~\cite{skorstengaard_reasoning_2017}.
Our new formulation starts from the premise that security results for a calling convention should be reusable as part of a larger proof of a secure compiler.
To accomodate this, we define a second operational semantics for our target language, which has a native well-bracketed stack and primitive ways to do calls and returns.
This well-behaved semantics guarantees WBCF and LSE natively for components using our calling convention.
As such, these components can be sure that they will only ever interact with well-behaved other components that respect our desirable properties.
To express security of our calling convention, we then show that considering the same components in the original semantics, does not give adversaries additional ways to interact with them. 
More formally, we show that mapping a component in the well-behaved semantics to the same component in the original semantics is fully abstract~\cite{abadi_protection_1999}, i.e. that components are indistinguishable to arbitrary adversaries in the well-behaved language iff they are indistinguishable to arbitrary adversaries in the original language.

@@

\section{Linear Stack and Return Capabilities}

\begin{figure}
  \centering
  \begin{subfigure}{0.49\linewidth}
    \centering
    \begin{tikzpicture}[scale=.5, every node={scale=.5}]
      % recurrent parts
      \stdstackstart[13]
      \inactsf{(0,2)}{(4.5,4)} {\footnotesize Trusted\\ \footnotesize stack frame 1}
      \inactadv{(0,4)}{(4.5,7)} {\footnotesize Adv. stack frame 1}
      \inactsf{(0,7)}{(4.5,9)} {\footnotesize Trusted \\ \footnotesize stack frame 2}
      \actadv{(0,9)}{(4.5,12)} {\footnotesize Adv. stack frame 2}

      % Stack pointer 1
      \begin{scope}
        \clip (4.6,4) rectangle (9,13);
        \capbracebot{(4.6,4)}{(4.6,13.5)}{adv. stack\\\footnotesize cap. 1}
      \end{scope}
      Stack pointer 2
      \begin{scope}
        \clip (4.8,4) rectangle (9,13.5);
        \capbrace{(5.2,9)}{(5.2,14)}{adv. stack\\\footnotesize cap. 2}
      \end{scope}

      \draw[red,thick,->] (4.5,10) to[out=0,in=50] node[midway,right] {} (7.5,7.3);
      \draw[red,thick,->] (7,7.3) to[out=115,in=0] node[midway,right] {} (4.5,8);
    \end{tikzpicture}
    \caption{An adversary uses a previous stack frame's stack pointer.}
    \label{fig:stack-ptr-abuse}
  \end{subfigure}
  \begin{subfigure}{0.49\linewidth}
    \centering
    \begin{tikzpicture}[scale=.5, every node={scale=.5}]
      % recurrent parts
      \stdstackstart[13]
      \inactsf{(0,2)}{(4.5,4)} {\footnotesize Trusted\\ \footnotesize stack frame 1}
      \inactadv{(0,4)}{(4.5,7)} {\footnotesize Adv. stack frame 1}
      \inactsf{(0,7)}{(4.5,9)} {\footnotesize Trusted \\ \footnotesize stack frame 2}
      \actadv{(0,9)}{(4.5,12)} {\footnotesize Adv. stack frame 2}
      \fill[red,draw=red,opacity=.4] (-.1,3.9) rectangle (4.6,9.1);
      \draw[red,thick,->] (4.5,10.5) to[out=0,in=0] node[midway,right,align=center] {return\\ to wrong\\ return ptr} (4.5,3);
    \end{tikzpicture}
    \caption{An adversary jumps to a previous stack frame's stack pointer.}
    \label{fig:ret-ptr-abuse}
  \end{subfigure}

  \caption{Possible ways to abuse stack and return capabilities}
  \label{fig:stack-ret-ptr-abuse}
\end{figure}

Our StackTokens calling convention is based on a traditional single stack, shared between all components.
As we are on a capability machine, it is natural to add some extra protection to stack and return pointers.
First, we replace stack pointers with stack capabilities.
When a new stack frame is created, the caller provisions it with a stack capability, restricted to the appropriate range, i.e.\ it does not cover the caller's stack frame.
Return pointers, on the other hand, are replaced by a sealed pair of return capabilities.
They form an opaque closure that the callee can only jump to, and the caller's data becomes available to the caller's return code.

While the above adds extra protection, it is important to understand that the above is not sufficient for enforcing WBCF and LSE.
While adversaries only get access to capabilities that they are allowed to use during the stack frame, they have ways to store these capabilities and use them when they are no longer allowed to do so.
Figure~\ref{fig:stack-ret-ptr-abuse} illustrates two examples of this.
The figure on the left (Figure~\ref{fig:stack-ptr-abuse}) blabla

\begin{itemize}
\item informally explain how an adversary may try to abuse stack and return caps
\item informally explain how we prevent this using linear capabilities
\item use the tikz pictures from the PriSC presentation to explain all of this
\end{itemize}
\section{A Capability Machine with Sealing and Linear Capabilities}
\begin{itemize}
\item present our target language and its operational semantics (excerpts)
\item mention roughly what components look like
\end{itemize}
\section{Formulating Security with a Fully Abstract Overlay Semantics}
\begin{itemize}
\item present our source language, its operational semantics (excerpts)
\item mention our assumption of reasonability
\item present the full abstraction theorem.
\item sketch high-level structure of the proof
\end{itemize}
\section{Discussion}
\begin{itemize}
\item mention that tail calls are supported through xcall
\item explain how fully abstract overlay semantics could form one pass of a verified secure compiler.
\end{itemize}
\section{Related Work}
\begin{itemize}
\item CFI and friends
\item our ESOP 2018 paper \citep{skorstengaard_reasoning_2017}
\item undefined behavior and full abstraction \citep{juglaret_beyond_2016}
\end{itemize}
\bibliography{references}

### 2018-07-01 fully-abstract-comp/popl2019/preamble-illu.tex
Some work on explaining the remaining problems with stack and return capabilities.

% Active adversary
\newcommand{\actadv}[4][]{
  \ifthenelse{\equal{#1}{}}{
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw ($#2 + (2.25,1.5)$) node[devil,mirrored,minimum size=1cm] {};
    \draw[fill=white, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw #2 rectangle #3 node[pos=.5,align=center] {\footnotesize #4};
  }{
    \onlyenv<#1>
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw ($#2 + (3,1.5)$) node[devil,mirrored,minimum size=1cm] {};
    \draw[fill=white, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw #2 rectangle #3 node[pos=.5] {\footnotesize #4};
    \endonlyenv
  }
}

% Inactive adversary
\newcommand{\inactadv}[4][]{
  \ifthenelse{\equal{#1}{}}{
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw ($#2 + (2.25,1.5)$) node[devil,mirrored,minimum size=1cm] {};
    \draw[fill=white, fill opacity=0.4] #2 rectangle #3 node[pos=.5] {};
    \draw[fill=gray, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw #2 rectangle #3 node[pos=.5] {\footnotesize #4};
  }{
    \onlyenv<#1>
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw ($#2 + (3,1.5)$) node[devil,mirrored,minimum size=1cm] {};
    \draw[fill=white, fill opacity=0.4] #2 rectangle #3 node[pos=.5] {};
    \draw[fill=gray, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw #2 rectangle #3 node[pos=.5] {\footnotesize #4};
    \endonlyenv
  }
}

% Active stack frame
\newcommand{\actsf}[4][]{
  \ifthenelse{\equal{#1}{}}{
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {\footnotesize #4};
  }{
   \draw<#1>[fill=white] #2 rectangle #3 node[pos=.5] {#4};
 }
}

% Inactive stack frame
\newcommand{\inactsf}[4][]{
  \ifthenelse{\equal{#1}{}}{
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw[fill=gray, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw #2 rectangle #3 node[pos=.5] {\footnotesize #4};
  }{
    \draw<#1>[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw<#1>[fill=gray, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw<#1> #2 rectangle #3 node[pos=.5] {\footnotesize #4};
  }
}

\newcommand{\capbrace}[4][sp1]{
  \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
  #2 -- #3 node (#1) [black,midway,xshift=0.8cm] {\footnotesize #4};
}

\newcommand{\capbracebot}[4][sp1]{
  \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt,aspect=0.25},yshift=0pt]
  #2 -- #3 node (#1) [black,near start,xshift=0.8cm] {\footnotesize #4};
}

\newcommand{\capbracebotred}[4][sp1]{
  \draw [decorate,red,decoration={brace,amplitude=10pt,mirror,raise=4pt,aspect=0.25},yshift=0pt]
  #2 -- #3 node (#1) [red,near start,xshift=0.8cm] {\footnotesize #4};
}

\newcommand{\stdstackstart}{
  \scope
    \clip (-.1,-.1) rectangle (4.6,11.1);
    \fill[fill=white] (0,0) rectangle (4.5,11);
    \draw (0,0) -- (0,11);
    \draw (4.5,0) -- (4.5,11);
    \draw[fill=gray!50] (0,-.5) rectangle (4.5,2)
    node[pos=.5,color=black,align=center,yshift=0.1cm] {\footnotesize Higher
      stack\\\footnotesize frames...};
  \endscope
  % \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);
  \draw (-1.5,0) node {};
  \draw (-1.5,15) node {};
}

>>>>

% Active adversary
\newcommand{\actadv}[4][]{
  \ifthenelse{\equal{#1}{}}{
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw ($#2 + (2.25,1.5)$) node[devil,mirrored,minimum size=1cm] {};
    \draw[fill=white, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw #2 rectangle #3 node[pos=.5,align=center] {\footnotesize #4};
  }{
    \onlyenv<#1>
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw ($#2 + (3,1.5)$) node[devil,mirrored,minimum size=1cm] {};
    \draw[fill=white, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw #2 rectangle #3 node[pos=.5,align=center] {\footnotesize #4};
    \endonlyenv
  }
}

% Inactive adversary
\newcommand{\inactadv}[4][]{
  \ifthenelse{\equal{#1}{}}{
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw ($#2 + (2.25,1.5)$) node[devil,mirrored,minimum size=1cm] {};
    \draw[fill=white, fill opacity=0.4] #2 rectangle #3 node[pos=.5] {};
    \draw[fill=gray, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw #2 rectangle #3 node[pos=.5,align=center] {\footnotesize #4};
  }{
    \onlyenv<#1>
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw ($#2 + (3,1.5)$) node[devil,mirrored,minimum size=1cm] {};
    \draw[fill=white, fill opacity=0.4] #2 rectangle #3 node[pos=.5] {};
    \draw[fill=gray, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw #2 rectangle #3 node[pos=.5,align=center] {\footnotesize #4};
    \endonlyenv
  }
}

% Active stack frame
\newcommand{\actsf}[4][]{
  \ifthenelse{\equal{#1}{}}{
    \draw[fill=white] #2 rectangle #3 node[pos=.5,align=center] {\footnotesize #4};
  }{
   \draw<#1>[fill=white] #2 rectangle #3 node[pos=.5,align=center] {#4};
 }
}

% Inactive stack frame
\newcommand{\inactsf}[4][]{
  \ifthenelse{\equal{#1}{}}{
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw[fill=gray, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw #2 rectangle #3 node[pos=.5,align=center] {\footnotesize #4};
  }{
    \draw<#1>[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw<#1>[fill=gray, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw<#1> #2 rectangle #3 node[pos=.5,align=center] {\footnotesize #4};
  }
}

\newcommand{\capbrace}[4][sp1]{
  \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
  #2 -- #3 node (#1) [black,midway,xshift=1cm,align=center] {\footnotesize #4};
}

\newcommand{\capbracebot}[4][sp1]{
  \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt,aspect=0.25},yshift=0pt]
  #2 -- #3 node (#1) [black,near start,xshift=1cm,align=center] {\footnotesize #4};
}

\newcommand{\capbracebotred}[4][sp1]{
  \draw [decorate,red,decoration={brace,amplitude=10pt,mirror,raise=4pt,aspect=0.25},yshift=0pt]
  #2 -- #3 node (#1) [red,near start,xshift=1cm,align=center] {\footnotesize #4};
}

\newcommand{\stdstackstart}[1][11]{
  \scope
    \clip (-.1,-.1) rectangle (4.6,#1 + .1);
    \fill[fill=white] (0,0) rectangle (4.5,#1);
    \draw (0,0) -- (0,#1);
    \draw (4.5,0) -- (4.5,#1);
    \draw[fill=gray!50] (0,-.5) rectangle (4.5,2)
    node[pos=.5,color=black,align=center,yshift=0.1cm] {\footnotesize Higher
      stack\\\footnotesize frames...};
  \endscope
  % \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);
  \draw (-1.5,0) node {};
  \draw (-1.5,#1 + 1) node {};
}
